Learning constraints in spreadsheets and tabular data

### 2016-06-12 Paper/ECAI/tabular.tex
Final read & edits [Samuel]

\newcommand{\sg}{B}
\newcommand{\sbl}[1]{\ensuremath{B_{\textit{#1}}}}
\newcommand{\bsbl}[1]{\ensuremath{\mathbf{B_{\textit{#1}}}}}

@@

\begin{abstract}
%Spreadsheets, comma separated value files and other tabular data representations are in wide use today. However, the correct use and maintenance of functions in and between the tabular data can be error-prone and overwhelming. In this work, we investigate the automatic discovery of constraints (functions and relations) from raw tabular data. Our method takes inspiration form inductive logic programming and constraint satisfiability. We represent common spreadsheet functions as predicates with arguments that must satisfy number of constraints over the arguments. This allows generic constraint satisfaction techniques to be used to find all such satisfying predicates/functions within the rows, columns and submatrices of the tables. We show the effectiveness and accuracy of this approach on a number of spreadsheets from varying sources.
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, modeling, maintaining and discovering formulas in tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic discovery of constraints (functions and relations) in raw tabular data.
We see multiple promising applications for this technique, e.g. in rediscovering constraints, auto-completion and error checking.
Our method takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
Common spreadsheet functions are represented as predicates whose arguments must satisfy a number of constraints.
Constraint learning techniques are used to identify the predicates that hold between blocks of rows and columns of the tables.
%New types of constraints can be easily added to the system by specifying them in a declarative language.
We show that our approach is able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

@@

In this paper,  we investigate whether learning techniques can be used to infer constraints (formulas and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where header names such as \textbf{Total} and \textbf{Rank} already suggest the usage of spreadsheet operations such as \textit{SUM} or \textit{RANK}.
Looking at the first row in Table~$T_2$ and the data in Table~$T_1$, it is clear that the values in this row were computed by summing the values in the corresponding columns (\textbf{1st Quarter} through \textbf{Total}) of $T_1$.
Examining $T_1$ also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by summing the rows of the previous four columns.
This provides a flavor of how this problem is different from standard mining and learning settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple tables with relationships between them (e.g. foreign-keys). And on the other hand, the data is most often mixed textual and numeric. %, since people tend to use spreadsheets for financial and accounting computations.
Hence understanding and learning constraints in tabular data is a novel and challenging problem for machine learning.

@@

The approach that we introduce borrows techniques from logical and relational learning \cite{luc_book}, where the discovery of clausal constraints has been studied in inductive logic programming \cite{claudien,lallouet}; and from constraint learning,  where several approaches to learning sets of constraints have been developed in a constraint programming setting \cite{Quacq,Conacq,modelseeker};
and from mining in databases, where inductive techniques have been applied to discover constraints (such as functional and multi-valued dependencies) in relational databases \cite{savnik}.  It also builds on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from
very few examples.
Our approach contrasts with these in that it focuses
on learning both column- and row-constraints on integer, floating point as well as textual data.

This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach. Section \ref{sec:approach} presents the problem statement and the approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents a more detailed discussion of the related work. Section \ref{sec:conclusions} provides conclusions.

@@

  \begin{subfigure}{.66\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo2.png}
  \end{center}
  \vspace{-10pt}
  \caption{
    Example spreadsheet (black words and numbers only).
    Green background indicates headerless tables, dark borders indicate maximal type-consistent blocks.
    Most tables only contain type-consistent columns while table~$T_2$ also contains type-consistent rows.
    However, due to the location of the header of~$T_2$ we can infer that it consists of row data.
    This example was created by combining several Excel sheets from several exercise sessions.
    % I guess you want to save space by having everything in one figure, but it might be useful to have a smaller figure on the first page just to explain this example in a bit more detail (or use a smaller example here); it's quite overwhelming now}
  }
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.34\textwidth}
  {\footnotesize
    \begin{align*}
      % Table 1
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
%      &~\ecperm{\range{T_{1}}{\rangeall}{1}}, \ecperm{\range{T_{1}}{\rangeall}{8}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{5}}^*\\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{6}}^* \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{10}}^*\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{3}}^*\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{4}}^* \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{3}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{3}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{3}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{3}}{\rangeall}{2}} \\
%
	  % Table 2
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}},  \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
	  % Table 4
      &~\ecsumc{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\range{T_{4}}{\rangeall}{4} = PREV(\range{T_{4}}{\rangeall}{4}) + \range{T_{4}}{\rangeall}{2} - \range{T_{4}}{\rangeall}{3} \\
%
    % Table 5
      &~\eclookup{\range{T_{5}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{3}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}^* \\
%
      &~\eclookup{\range{T_{5}}{\rangeall}{3}}{\range{T_{5}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}
    \end{align*}}
  \vspace{-18pt}
  \caption{Constraints learned for the example (left), except \textit{ALLDIFFERENT} (16), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).
  Constraints marked with $*$ were not present in the original document}
  \label{fig:sol_example}
\end{subfigure}
  \caption{Running example}
\end{figure*}

@@

A row or a column is \textbf{type-consistent} if all cells in that row or column are of the same base type, i.e. numeric or textual.
We will use notation $T[a,{:}]$ to refer to the $a$-th row of table $T$ and similarly $T[{:},a]$ for the $a$-th column.
For example, in Figure~\ref{fig:main_example}, $T_1[1,:] = [1,2,3,4]$ and $T_5[:,1] = ['$Diana Coolen$', 5]$.
$T_5[:,1]$ is not type-consistent while $T_1[1,:]$ is.
%We will write \textbf{vector} to refer to a single row or column when its orientation does not matter.

The most important concept is that of a \textbf{block}. 
\begin{definition}
A \textbf{block} has to satisfy three conditions: 1) it contains only entire rows or entire columns of a single headerless table 2) it is contiguous and 3) it is type-consistent.
The rows or columns have to be contiguous in the original table meaning that they must visually form a block in the table; and each of the rows/columns has to be of the same type. 
If it contains only rows we say it has \textit{row-orientation}, if only columns, \textit{column-orientation}. 
\end{definition}

@@

\begin{example}
Consider the (headerless) table~$T_1$ in Figure~\ref{fig:main_example}.
Its rows are not type consistent (i.e. they contain both numeric and textual data).
However, the table can be partitioned into five column-oriented blocks $B_1, B_2, B_3, B_4, B_5$, as shown in the figure ($B_1 = \range{T_1}{\rangeall}{1}$, $B_2 = \range{T_1}{\rangeall}{2}$, $B_3 = \range{T_1}{\rangeall}{\rangeto{3}{8}}$, \dots).
% {\small
% B_1 = \range{T_1}{\rangeall}{1},
% B_2 = \range{T_1}{\rangeall}{2},
% B_3 = \range{T_1}{\rangeall}{\rangeto{3}{8}},
% B_4 = \range{T_1}{\rangeall}{9},
% B_5 = \range{T_1}{\rangeall}{\rangeto{10}{11}}
\end{example}

@@

\begin{table*}[!h]
\caption{
  Overview of constraint templates implemented in \sname.
  Subgroups in normal font have to be vectors, whereas subgroups in \textbf{bold} may contain more than one vector;
  $\discrete(x)$ is a shortcut for: $\textual(x) \lor \numeric(x)$.
  Templates marked with $*$ are \textit{structural}, i.e. they are not functional and cannot be implemented in most spreadsheet software.
  Templates marked with $\dagger$ are \textit{aggregate} templates, the table shows \textit{SUM} but \sname also supports \textit{MAX}, \textit{MIN}, \textit{AVERAGE}, \textit{PRODUCT} and \textit{COUNT}.
  %{\small\luc{maybe say s.th. about some fancy constraints: example of a more complex one in the text (what does conditional mean, aggregate ... )}}
}
\label{table:constraints}
  {\centering
  \begin{tabularx}{\textwidth}{l X X}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\ \hline \hline
    $\ecalldiff{\sg_x}^*$
      & $\discrete(\sg_x)$

      & All values in $\sg_x$ are different. $i \neq j$: $\sg_x[i] \neq \sg_x[j]$
      \\[\PADSEP] \hline

    $\ecperm{\sg_x}^*$
      & $\numeric(\sg_{x})$, $\ecalldiff{\sg_{x}}$

      & The values in $\sg_{x}$ are a permutation of the numbers $1$ through $\plength(\sg_{x})$.
      \\ \hline

    \ecseries{\sg_x}
      & $\integer(\sg_{x})$ and $\ecperm{\sg_{x}}$

      & $\sg_{x}[1] = 1$ and $\sg_{x}[i] = \sg_{x}[i - 1] + 1$.
      \\[\PADSEP] \hline

    $\ecfkey{\sg_{fk}}{\sg_{pk}}^*$
      & $\sg_{\mathit{fk}}$ and~$\sg_{pk}$ are $\discrete$; they must have different $\ptable$s but the same $\ptype$; and $\ecalldiff{\sbl{pk}}$

      & Every value in~$\sg_{fk}$ also exist in~$\sg_{pk}$ \\[\PADSEP] \hline

@@

    \eclookupprod{\sbl{r}}{\sbl{1}}{\sbl{fk}}{\sbl{pk}}{\sbl{val}}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{fk}\}$ are $\numeric$, arguments $\{\sg_{pk}, \sg_{val}\}$ are $\discrete$ and within both sets all arguments have the same \plength, \ptable and \por; also \ecfkey{\sg_{fk}}{\sg_{pk}}.

@@

    \ecprod{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{2}\}$ are all $\numeric$ and have the same $\plength$

      & $\sg_{r}[i] = \sg_{1}[i] \times \sg_{2}[i]$.
      \\[\PADSEP] \hline

    \ecdiff{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_{r}, \sg_{1}, \sg_{2}\}$ are all $\numeric$ and have the same $\plength$ and $ \por$

      & $\sg_{r}[i] = \sg_{1}[i] - \sg_{2}[i]$.
      \\[\PADSEP] \hline

    $\eccalc{\sg_r}{\frac{\sg_1 - \sg_2}{\sg_2}}$
      & \textit{Same as \ecdiff{\sg_r}{\sg_1}{\sg_2}}

      & $\sg_{r}[i] = (\sg_{1}[i] - \sg_{2}[i]) / \sg_{2}[i]$.
      \\[\PADSEP] \hline

    \ecproj{\sg_r}{\mathbf{\sg_x}}
      & Arguments $\{\sg_{r}, \mathbf{\sg_x}\}$ all have the same $\plength$, $\por$, $\ptable$ and $\ptype$; $\mathbf{\sg_x}$ contains at least~2 vectors; and $\sg_r = \mathit{SUM}_{\por(\mathbf{\sg_x})}(\mathbf{\sg_x})$

      & At every position~$i$ in $1$ through $\plength(\sg_{r})$ there is exactly one vector~$v$ in $\mathbf{\sg_x}$ such that $v[i]$ is a non-blank value, then $v[i] = \sg_{r}[i]$.
      \\[\PADSEP] \hline

    \ecrank{\sg_r}{\sg_x}
      & $\integer(\sg_{r})$; $\numeric(\sg_{x})$; and $\plength(\sg_{r}) = \plength(\sg_{x})$

      & The values in $\sg_{r}$ represent the rank (from largest to smallest) of the values in $\sg_{x}$ (including ties)
      \\[\PADSEP] \hline

    \ectotal{\sg_r}{\sg_{pos}}{\sg_{neg}}
      & Arguments $\{\sg_{r}, \sg_{pos}, \sg_{neg}\}$ are all $\numeric$ and all have the same $\plength$, which is at least $2$

      & $\sg_{r}[i] = \sg_{r}[i - 1] + \sg_{pos}[i] - \sg_{neg}[i]$.
      \\[\PADSEP] \hline

    $\ecsumr{\sg_r}{\mathbf{\sg_x}}^\dagger$
      & $\sg_r$ and $\mathbf{\sg_x}$ are $\numeric$; $\pcols(\mathbf{\sg_x}) \geq 2$; and $\prows(\mathbf{\sg_x}) = \plength(\sg_r)$

      & $B_r[i] = \sum_{j = 1}^{\pcols(\bsbl{x})} \mathit{row}(i, \bsbl{x})[j]$
      \\[\PADSEP] \hline

    $\ecsumc{\sg_r}{\mathbf{\sg_x}}^\dagger$
      & $\sg_r$ and $\mathbf{\sg_x}$ are $\numeric$; $\prows(\mathbf{\sg_x}) \geq 2$; and $\pcols(\mathbf{\sg_x}) = \plength(\sg_r)$

      & $B_r[i] = \sum_{j = 1}^{\prows(\bsbl{x})} \mathit{column}(i, \bsbl{x})[j]$
      \\[\PADSEP] \hline

    $\ecsumif{\sg_r}{\sg_{fk}}{\sg_{pk}}{\sg_{val}}^\dagger$
      & $\sg_{fk}, \sg_{pk}$ are $\discrete$; $\sg_{r}, \sg_{val}$ are $\numeric$; within the sets $\{\sg_{val}, \sg_{fk}\}$ and $\{\sg_{pk}, \sg_{r}\}$ arguments have the same $\plength$ and $\por$; $\sg_{fk}$ and $\sg_{val}$ have the same $\ptable$; $\sg_{fk}$ and $\sg_{pk}$ must have different $\ptable$s but the same $\ptype$; and \ecalldiff{\sg_{pk}}

      & \[ B_r[i] = \sum_{j=1}^{\plength(B_{\textit{val}})} \begin{cases}
          \sg_{val}[j] & \text{if } \sg_{fk}[j] = \sg_{pk}[i] \\
          0 & \text{otherwise}
        \end{cases}
      \] \\[\PADSEP] \hline

    \ecsumprod{\sg_r}{\sg_1}{\sg_2}
      & Arguments $\{\sg_r, \sg_1, \sg_2\}$ are $\numeric$; $\plength(\sg_{1}) = \plength(\sg_{2}) \geq 2$; and $\prows(\sg_{r}) = \pcols(\sg_{r}) = 1$

      & $\sg_{r}[1] = \sum_{j = 1}^{\plength(\sg_{1})} \sg_{1}[j] \times \sg_{2}[j]$.
  \end{tabularx}}

@@

\begin{itemize}
  \item \CName: $\ecsumr{\sg_r}{\mathbf{\sg_x}}$, for arguments $\sg_r$ and $\mathbf{\sg_x}$.

  \item \CSignature: $\sg_r$ has to be a single vector ($\textit{size}=1$) while $\mathbf{\sg_x}$ can be a block ($\textit{size}>=1$), which can be derived from the use of a normal or \textbf{bold} font.
  Both blocks have to be numeric.
  This constraint is orientation-specific, so it requires that the number of rows in $\mathbf{\sg_x}$ equals the length of $\sg_r$.
  Moreover, we add to the bias that the number of columns to sum over is at least~$2$.

>>>>

\newcommand{\sbs}{B}
\newcommand{\sbl}[1]{\ensuremath{\sbs_{\textit{#1}}}}
\newcommand{\ssbl}[1]{\ensuremath{\sbs'_{\textit{#1}}}}
\newcommand{\bsbl}[1]{\ensuremath{\mathbf{\sbs_{\textit{#1}}}}}

@@

\begin{abstract}
Spreadsheets, CSV (comma separated value) files and other tabular data representations are in wide use today.
However, modeling, maintaining and discovering formulas in tabular data and spreadsheets can be time consuming and error-prone.
In this work, we investigate the automatic discovery of constraints (functions and relations) in raw tabular data.
We see multiple promising applications for this technique, e.g. in rediscovering constraints, auto-completion and error checking.
Our method takes inspiration from inductive logic programming, constraint learning and constraint satisfaction.
Common spreadsheet functions are represented as predicates whose arguments must satisfy a number of constraints.
Constraint learning techniques are used to identify the predicates that hold between blocks of rows and columns of the tables.
We show that our approach is able to accurately discover constraints in spreadsheets from various sources.
\end{abstract}

@@

In this paper,  we investigate whether learning techniques can be used to infer constraints (formulas and other relations) from raw spreadsheet data.
This is a new and unconventional machine learning problem.
Consider the example in Figure~\ref{fig:main_example}, where header names such as \textbf{Total} and \textbf{Rank} already suggest the usage of spreadsheet operations such as \textit{SUM} or \textit{RANK}.
Looking at the first row in table~$T_2$ and the data in table~$T_1$, it is clear that the values in this row were computed by summing the values in the corresponding columns (\textbf{1st Quarter} through \textbf{Total}) of $T_1$.
Examining $T_1$ also shows that computations are not only performed column-wise but row-wise as well: the cells in the column \textbf{Total} are obtained by summing the rows of the previous four columns.
This provides a flavor of how this problem is different from standard mining and learning settings, where the data is just in rows and variables are in columns. Here everything is mixed. The data is relational on the one hand, since we have multiple tables with relationships between them (e.g. foreign-keys). On the other hand, the data is most often mixed, e.g. textual and numeric. %, since people tend to use spreadsheets for financial and accounting computations.
Therefore, understanding and learning constraints in tabular data is a novel and challenging problem for machine learning.

@@

The approach that we introduce borrows techniques from logical and relational learning \cite{luc_book}, where the discovery of clausal constraints has been studied in inductive logic programming \cite{claudien,lallouet}; and from constraint learning,  where several approaches to learning sets of constraints have been developed in a constraint programming setting \cite{Quacq,Conacq,modelseeker};
and from mining in databases, where inductive techniques have been applied to discover constraints (such as functional and multi-valued dependencies) in relational databases \cite{savnik}.  It also builds on work on program synthesis, in particular, on Flashfill \cite{flashfill}, where the definition of a function (over textual cells only) is learned in spreadsheet data from
very few examples.
Our approach contrasts with these in that it focuses on learning both column- and row-constraints across tables on integer, floating point as well as textual data.

This paper is organized as follows.
Section \ref{sec:formalization} introduces concepts relevant to our approach and our problem statement. Section \ref{sec:approach} presents our approach. Section \ref{sec:evaluation} presents the evaluation of the approach. Section \ref{sec:applications} shows how our system can be used for applications. Section \ref{sec:related_work} presents a more detailed discussion of the related work. Section \ref{sec:conclusions} provides conclusions.

@@

  \begin{subfigure}{.66\textwidth}
  \begin{center}
    \includegraphics[width=1\textwidth]{figures/Demo2.png}
  \end{center}
  \vspace{-10pt}
  \caption{
    Example spreadsheet (black words and numbers only).
    Green background indicates headerless tables, dark borders indicate maximal type-consistent blocks.
    Most tables only contain type-consistent columns while table~$T_2$ also contains type-consistent rows.
    However, due to the location of the header of~$T_2$ we can infer that it consists of row data.
    This example was created by combining several Excel sheets from several exercise sessions based on~\cite{excel_book}.
    % I guess you want to save space by having everything in one figure, but it might be useful to have a smaller figure on the first page just to explain this example in a bit more detail (or use a smaller example here); it's quite overwhelming now}
  }
  \label{fig:main_example}
\end{subfigure}
\hfill
\begin{subfigure}{.34\textwidth}
  {\footnotesize
    \begin{align*}
      % Table 1
      &~\ecseries{\range{T_{1}}{\rangeall}{1}} \\
%
%      &~\ecperm{\range{T_{1}}{\rangeall}{1}}, \ecperm{\range{T_{1}}{\rangeall}{8}} \\
%
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{5}}^*\\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{6}}^* \\
      &~\ecrank{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{10}}^*\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{3}}^*\\
      &~\ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{4}}^* \\
%
      &~\ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\ecaggif{\textit{SUM}}{\range{T_{1}}{\rangeall}{10}}{\range{T_{3}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{3}}{\rangeall}{2}} \\
%
      &~\ecaggif{\textit{MAX}}{\range{T_{1}}{\rangeall}{11}}{\range{T_{3}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{3}}{\rangeall}{2}} \\
%
	  % Table 2
      &~\ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{AVERAGE}}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      &~\ecaggc{\textit{MAX}}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}},  \\
%
      &~\ecaggc{\textit{MIN}}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
	  % Table 4
      &~\ecsumc{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      &~\range{T_{4}}{\rangeall}{4} = PREV(\range{T_{4}}{\rangeall}{4}) + \range{T_{4}}{\rangeall}{2} - \range{T_{4}}{\rangeall}{3} \\
%
    % Table 5
      &~\eclookup{\range{T_{5}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{3}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}^* \\
%
      &~\eclookup{\range{T_{5}}{\rangeall}{3}}{\range{T_{5}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}
    \end{align*}}
  \vspace{-18pt}
  \caption{Constraints learned for the example (left), except for \textit{ALLDIFFERENT} (18), \textit{PERMUTATION} (2) and \textit{FOREIGNKEY} (5).
  Constraints marked with $*$ were not present in the original document}
  \label{fig:sol_example}
\end{subfigure}
  \caption{Running example}
\end{figure*}

@@

A row or a column is \textbf{type-consistent} if all cells in that row or column are of the same base type, i.e. numeric or textual.
We will use notation $T[a,{:}]$ to refer to the $a$-th row of table $T$ and similarly $T[{:},a]$ for the $a$-th column.
For example, in Figure~\ref{fig:main_example}, $T_1[1,:] = [1,2,3,4]$ and $T_3[:,1] = ['$Diana Coolen$', 5]$.
$T_3[1, :]$ is not type-consistent while $T_1[:, 1]$ is.
%We will write \textbf{vector} to refer to a single row or column when its orientation does not matter.

The most important concept is that of a \textbf{block}. 
\begin{definition}
A \textbf{block} has to satisfy three conditions: 1)~it contains only entire rows or entire columns of a single headerless table; 2)~it is contiguous; and 3)~it is type-consistent.
The rows or columns have to be contiguous in the original table meaning that they must visually form a block in the table; and each of the rows/columns has to be of the same type. 
If a block contains only rows we say it has \textit{row-orientation}, if only columns, \textit{column-orientation}. 
\end{definition}

@@

\begin{example}
Consider the (headerless) table~$T_1$ in Figure~\ref{fig:main_example}.
Its rows are not type consistent (i.e. they contain both numeric and textual data).
However, the table can be partitioned into five column-oriented blocks $\sbl{1}, \sbl{2}, \sbl{3}, \sbl{4}, \sbl{5}$, as shown in the figure ($\sbl{1} = \range{T_1}{\rangeall}{1}$, $\sbl{2} = \range{T_1}{\rangeall}{2}$, $\sbl{3} = \range{T_1}{\rangeall}{\rangeto{3}{8}}$, \dots).
\end{example}

@@

\begin{table*}[!h]
\caption{
  Overview of constraint templates implemented in \sname.
  Subgroups in normal font have to be vectors, whereas subgroups in \textbf{bold} may contain more than one vector;
  $\discrete(x)$ is a shortcut for: $\textual(x) \lor \numeric(x)$.
  Templates marked with $*$ are \textit{structural}, i.e. they are not functional and cannot be implemented in most spreadsheet software.
  Templates marked with $\dagger$ are \textit{aggregate} templates, the table shows \textit{SUM} but \sname also supports \textit{MAX}, \textit{MIN}, \textit{AVERAGE}, \textit{PRODUCT} and \textit{COUNT}.
  %{\small\luc{maybe say s.th. about some fancy constraints: example of a more complex one in the text (what does conditional mean, aggregate ... )}}
}
\label{table:constraints}
  {\centering
  \begin{tabularx}{\textwidth}{l X X}
    \textbf{\CName} & \textbf{\CSignature} & \textbf{\CFunction}\\ \hline \hline
    $\ecalldiff{\sbl{x}}^*$
      & $\discrete(\sbl{x})$

      & All values in $\sbl{x}$ are different. $i \neq j$: $\sbl{x}[i] \neq \sbl{x}[j]$
      \\[\PADSEP] \hline

    $\ecperm{\sbl{x}}^*$
      & $\numeric(\sbl{x})$, $\ecalldiff{\sbl{x}}$

      & The values in $\sbl{x}$ are a permutation of the numbers $1$ through $\plength(\sbl{x})$.
      \\ \hline

    \ecseries{\sbl{x}}
      & $\integer(\sbl{x})$ and $\ecperm{\sbl{x}}$

      & $\sbl{x}[1] = 1$ and $\sbl{x}[i] = \sbl{x}[i - 1] + 1$.
      \\[\PADSEP] \hline

    $\ecfkey{\sbl{fk}}{\sbl{pk}}^*$
      & $\sbl{fk}$ and~$\sbl{pk}$ are $\discrete$; they must have different $\ptable$s but the same $\ptype$; and $\ecalldiff{\sbl{pk}}$

      & Every value in~$\sbl{fk}$ also exist in~$\sbl{pk}$ \\[\PADSEP] \hline

@@

    \eclookupprod{\sbl{r}}{\sbl{1}}{\sbl{fk}}{\sbl{pk}}{\sbl{val}}
      & Arguments $\{\sbl{r}, \sbl{1}, \sbl{fk}\}$ are $\numeric$, arguments $\{\sbl{pk}, \sbl{val}\}$ are $\discrete$ and within both sets all arguments have the same \plength, \ptable and \por; also \ecfkey{\sbl{fk}}{\sbl{pk}}.

@@

    \ecprod{\sbl{r}}{\sbl{1}}{\sbl{2}}
      & Arguments $\{\sbl{r}, \sbl{1}, \sbl{2}\}$ are all $\numeric$ and have the same $\plength$

      & $\sbl{r}[i] = \sbl{1}[i] \times \sbl{2}[i]$.
      \\[\PADSEP] \hline

    \ecdiff{\sbl{r}}{\sbl{1}}{\sbl{2}}
      & Arguments $\{\sbl{r}, \sbl{1}, \sbl{2}\}$ are all $\numeric$ and have the same $\plength$ and $ \por$

      & $\sbl{r}[i] = \sbl{1}[i] - \sbl{2}[i]$.
      \\[\PADSEP] \hline

    $\eccalc{\sbl{r}}{\frac{\sbl{1} - \sbl{2}}{\sbl{2}}}$
      & \textit{Same as \ecdiff{\sbl{r}}{\sbl{1}}{\sbl{2}}}

      & $\sbl{r}[i] = (\sbl{1}[i] - \sbl{2}[i]) / \sbl{2}[i]$.
      \\[\PADSEP] \hline

    \ecproj{\sbl{r}}{\bsbl{x}}
      & Arguments $\{\sbl{r}, \bsbl{x}\}$ all have the same $\plength$, $\por$, $\ptable$ and $\ptype$; $\bsbl{x}$ contains at least~2 vectors; and $\sbl{r} = \textit{SUM}_{\por(\bsbl{x})}(\bsbl{x})$

      & At every position~$i$ in $1$ through $\plength(\sbl{r})$ there is exactly one vector~$v$ in $\bsbl{x}$ such that $v[i]$ is a non-blank value, then $v[i] = \sbl{r}[i]$.
      \\[\PADSEP] \hline

    \ecrank{\sbl{r}}{\sbl{x}}
      & $\integer(\sbl{r})$; $\numeric(\sbl{x})$; and $\plength(\sbl{r}) = \plength(\sbl{x})$

      & The values in $\sbl{r}$ represent the rank (from largest to smallest) of the values in $\sbl{x}$ (including ties)
      \\[\PADSEP] \hline

    \ectotal{\sbl{r}}{\sbl{pos}}{\sbl{neg}}
      & Arguments $\{\sbl{r}, \sbl{pos}, \sbl{neg}\}$ are all $\numeric$ and all have the same $\plength$, which is at least $2$

      & $\sbl{r}[i] = \sbl{r}[i - 1] + \sbl{pos}[i] - \sbl{neg}[i]$.
      \\[\PADSEP] \hline

    $\ecsumr{\sbl{r}}{\bsbl{x}}^\dagger$
      & $\sbl{r}$ and $\bsbl{x}$ are $\numeric$; $\pcols(\bsbl{x}) \geq 2$; and $\prows(\bsbl{x}) = \plength(\sbl{r})$

      & $\sbl{r}[i] = \sum_{j = 1}^{\pcols(\bsbl{x})} \textit{row}(i, \bsbl{x})[j]$
      \\[\PADSEP] \hline

    $\ecsumc{\sbl{r}}{\bsbl{x}}^\dagger$
      & $\sbl{r}$ and $\bsbl{x}$ are $\numeric$; $\prows(\bsbl{x}) \geq 2$; and $\pcols(\bsbl{x}) = \plength(\sbl{r})$

      & $\sbl{r}[i] = \sum_{j = 1}^{\prows(\bsbl{x})} \textit{column}(i, \bsbl{x})[j]$
      \\[\PADSEP] \hline

    $\ecsumif{\sbl{r}}{\sbl{fk}}{\sbl{pk}}{\sbl{val}}^\dagger$
      & $\sbl{fk}, \sbl{pk}$ are $\discrete$; $\sbl{r}, \sbl{val}$ are $\numeric$; within the sets $\{\sbl{val}, \sbl{fk}\}$ and $\{\sbl{pk}, \sbl{r}\}$ arguments have the same $\plength$ and $\por$; $\sbl{fk}$ and $\sbl{val}$ have the same $\ptable$; $\sbl{fk}$ and $\sbl{pk}$ must have different $\ptable$s but the same $\ptype$; and \ecalldiff{\sbl{pk}}

      & \[ \sbl{r}[i] = \sum_{j=1}^{\plength(\sbl{val})} \begin{cases}
          \sbl{val}[j] & \text{if } \sbl{fk}[j] = \sbl{pk}[i] \\
          0 & \text{otherwise}
        \end{cases}
      \] \\[\PADSEP] \hline

    \ecsumprod{\sbl{r}}{\sbl{1}}{\sbl{2}}
      & Arguments $\{\sbl{r}, \sbl{1}, \sbl{2}\}$ are $\numeric$; $\plength(\sbl{1}) = \plength(\sbl{2}) \geq 2$; and $\prows(\sbl{r}) = \pcols(\sbl{r}) = 1$

      & $\sbl{r}[1] = \sum_{j = 1}^{\plength(\sbl{1})} \sbl{1}[j] \times \sbl{2}[j]$.
  \end{tabularx}}

@@

\begin{itemize}
  \item \CName: $\ecsumr{\sbl{r}}{\bsbl{x}}$, for arguments $\sbl{r}$ and $\bsbl{x}$.

  \item \CSignature: $\sbl{r}$ has to be a single vector ($\textit{size}=1$) while $\bsbl{x}$ can be a block ($\textit{size}>=1$), which can be derived from the use of a normal or \textbf{bold} font.
  Both blocks have to be numeric.
  This constraint is orientation-specific, so it requires that the number of rows in $\bsbl{x}$ equals the length of $\sbl{r}$.
  Moreover, we add to the bias that the number of columns to sum over is at least~$2$.
